---
title: "ENVS-193DS_final"
author: "Celine Park, Lauren Stiles"
date: 06-2023
format:
  html:
    toc: true
    toc-location: left
    code-fold: true
    theme: yeti
execute: 
  message: false
  warning: false 
---


#load libraries: 
```{r message = FALSE, warning = FALSE}
library(MASS)
library(tidyverse)
library(here) 
library(janitor) 
library(readxl)
library(naniar)
library(car)
library(flextable)
```

## **Problem 1**

# Introduction 

Small scale biotic disturbances can influence species compositions in plant communities (Bullock, 2000). In the Chihuahuan desert, kangaroo rat mounds can be classified as this type of small scale disturbance (Koontz and Simpson, 2010). Previous studies comparing landscapes containing kangaroo rat mounds have found that annual plant cover is higher on top of mounds than in surrounding areas (Davidson and Lightfoot, 2008). Though the literature outlines difference between vegetation composition, the scientific community still lacks understanding of how kangaroo rat mounds might influence seed bank composition (Koontz and Simpson, 2010). In this analysis, we aim to investigate this potential relationship. The question we want to answer is: How does total seed number differ between kangaroo rat mound locations? Our null hypothesis is that kangaroo rat mounds do not influence seed bank composition and our alternative hypothesis is that kangaroo rat mounds do influence seed bank composition. 

# Methods 

We read in the data available [here](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-sev.208.102459) (Koontz, T. and H. Simpson. 2013). We cleaned the data by selecting the columns we are interested in and renaming the locations (base of mound, edge of grass, grass inter-space) for more straightforward interpretation. We then added up all of the seed counts in each category since we are comparing total counts. We chose to use a Chi-squared analysis to compare the total seed counts in each category since this kind of test is useful for analyzing count data. Our data meets each of the assumptions required to run a Chi-squared test. Our data represents counts and the samples are independent. Individual counts can only fall in one category at a time since each count corresponds to a specific sampling site. Finally, each expected value for each cell was greater than 5. We tested the null hypothesis that sampling location does not significantly affect total seed count. The alternative hypothesis is that sampling location does influence total counts. 

Read in data: 
```{r message = FALSE, warning = FALSE}
seed_dat <- read_csv(here("data/sev208_kratseedbank_20120213.csv"))
```

Clean data: 
```{r message = FALSE, warning = FALSE}
#create new data frame with kangaroo rat mound locations and seed numbers found 
seed_dat_clean <- seed_dat |> 
                  #select columns of interest
                  select(loc, seeds) |> 
                  #remove "D" from loc column becuase there is no data defintion on the EDI portal 
                  filter(loc != "D") |>
                  #rename columns 
                  mutate(loc = case_when(loc == "B" ~ "Base of mound", 
                                          loc == "E" ~ "Edge of grass", 
                                          loc == "I" ~ "Grass interspace")) |> 
                  group_by(loc) |> 
                  summarize(count = sum(seeds)) |> 
                  ungroup() |> 
                  column_to_rownames("loc")

#check for missing data 
gg_miss_var(seed_dat_clean) #no missing data 

#check how many sampling events for each location 
seed_dat_n <- seed_dat |> 
              select(loc, seeds) |> 
              mutate(loc = case_when(loc == "B" ~ "Base of mound", 
                                          loc == "E" ~ "Edge of grass", 
                                          loc == "I" ~ "Grass interspace")) |> 
              group_by(loc) |> 
              count(loc)
  
```


Run Chi-square Test: 
```{r}
seed_test <- chisq.test(seed_dat_clean)
seed_test
```




Visualization: 
```{r}
#reorganize seed_dat_clean so location is column name 
#rename columns 
seed_viz_dat <- seed_dat_clean |>  
                rownames_to_column("loc") 
  
#create bar graph to compare total seed counts at each location 
seed_viz <- ggplot(data = seed_viz_dat, aes(x = loc , y = count)) +
            geom_col(fill = "darkolivegreen4") + 
            labs(x = "Sampling Location", 
                 y = "Seed Count", 
                 title = "Total Number of Seeds Collected at each Sampling Location", 
                 caption = "Figure 1. Columns represent the total number of seeds collected.") + 
            theme_bw() + 
            #get rid of gap between bars and x axis 
            scale_y_continuous(limits = c(0, 7000),  expand = c(0, 0)) +
            #make plot look prettier
            theme(text = element_text(family = "Times New Roman"),
                  #change size and location of plot caption
                  plot.caption = element_text(size = 10, hjust = 0.5), 
                  #change size of plot title 
                  plot.title = element_text(size = 15), 
                  # getting rid of gridlines
                  panel.grid = element_blank()) 
                  #get rid of gap between bars and x axis)
            
seed_viz
```

**Table 1: Total seed count data by sampling location** Numbers in parentheses indicate the proportion of responses (the 6932 total seeds found at the base of kangaroo rat mounds make up 78 percent of the total seeds found at all locations, n = 1170). 
```{r}
# calculate proportions
seed_summary <- seed_dat_clean %>% 
#make location column from rownames 
 rownames_to_column("Location") %>% 
#find total number of seeds from all locations sampled and make new column 
  mutate(sum = sum(count)) %>% 
#ungroup to make sure that distance groups don't mess up downstream functions
  ungroup() %>% 
#calculate proportion of responses per living distance
 mutate(prop = count/sum) %>% 
#make a new column where counts and proportions are displayed together
  mutate(text = paste0(count, " (", round(prop, digits = 2), ")")) %>% 
#select columns of interest
  select(Location, text) %>% 
#making the data frame wider so that the columns are responses and rows are distance
  pivot_wider(names_from = "Location", values_from = "text") %>% 
#turning everything into a table ----
  flextable() %>% 
#changing the column names to look nicer
  set_header_labels(distance = "Living distance",
                    trails = "Trails",
                    dog_access = "Dog access",
                    wildlife_habitat = "Wildlife habitat") %>% 
#making the table fit the viewer window
  autofit()
  
seed_summary
```

# Results
Considering count data taken from the base of kangaroo rat mounds (n = 390), at the edge of  grass mounds Bouteloua eripoda (n = 390), and in the inter-space between Bouteloua eripoda grass mounds (n = 390), total seed counts differ significantly by sampling location (Table 1. X-squared = 8106 (2), p = < 2.2e-16, alpha = 0.05). We found that a large majority of seeds were collected at the base of the kangaroo rat mounds (78 percent), while significantly fewer seeds were found at locations beyond the kangaroo rat mounds (14 percent & 7 percent). This result indicates that there are more seeds in the seed bank at the kangaroo rat mounds than in the seed bank below nearby Bouteloua eripoda grasses. We can reject the null hypothesis that sampling location has no effect on total seed count and conclude that kangaroo rat mounds have an effect on seed bank composition at this study site. 


## **Bibliography**

Bullock, J. M. (2000) ‘Gaps and seedling colonization.’, CABI Books. CABI International. doi: 10.1079/9780851994321.0375.

Davidson, & Lightfoot, D. C. (2008). Burrowing rodents increase landscape heterogeneity in a desert grassland. Journal of Arid Environments, 72(7), 1133–1145. https://doi.org/10.1016/j.jaridenv.2007.12.015

Koontz, T. and H. Simpson. 2013. Effects of Kangaroo Rat Mounds on Seed Banks of Grass
and Shrublands at the Sevilleta National Wildlife Refuge, New Mexico (2001) ver 102459.
Environmental Data Initiative.

Koontz and Simpson 2010, “The composition of seed banks on kangaroo rat
(Dipodomysspectabilis) mounds in a Chihuahuan Desert grassland.” Journal of Arid Environments.
https://doi.org/10.1016/j.jaridenv.2010.03.008

## **Problem 2**

load libraries
```{r}
library(lme4)
# library(glmmTB) #not available 
library(DHARMa)
library(ggeffects)
library(lmtest)
library(broom)
library(MuMIn)
```


Read in data: 
```{r}
seedcount <-read_csv(here("data/knb-lter-nwt.297.1/seedcount.csv")) |> #reading in data 
            clean_names()  #adjusting column names to lowercase 
           

```

Visualizing missing data:
```{r}
gg_miss_var(seedcount)
```
Building models
```{r}
#linear model, this is wrong
seedcount1 <- lm(nr_seeds ~ total_nr_infl + plant_nr + tag_num, data = seedcount)

# general linear model with Poisson distribution
seedcount2 <- glm(nr_seeds ~ total_nr_infl + plant_nr + tag_num, data = seedcount, family = "poisson")
seedcoun2.a <- glm(nr_seeds ~ total_nr_infl + plant_nr + tag_num, data = seedcount, family = "poisson")

# general linear model with negative binomial distribution
seedcount3 <- glm.nb(nr_seeds ~ total_nr_infl + plant_nr + tag_num, data = seedcount)
#seedcount3.a <- glmmTB(nr_seeds ~ total_nr_infl + plant_nr + tag_num, data = seedcount, family = "nbinom2") don't have packages

#generalized linear model with Poisson distribution and random effect of site
seedcount4 <- glmer(nr_seeds ~ total_nr_infl + plant_nr + tag_num + (1|shrub_num), data = seedcount, family = "poisson")

#generalized linear model with negative binomial distribution and random effect of site
seedcount5 <- glmer.nb(nr_seeds ~ total_nr_infl + plant_nr + tag_num + (1|shrub_num), data = seedcount)



```


Selecting the distribution to use

```{r}
MuMIn::model.sel(seedcount1, seedcount2, seedcount3, seedcount4, seedcount5)
```
Model Summary
```{r}
seedcount2
```

```{r}
summary(seedcount2)
```
```{r}
confint(seedcount2)
```

```{r}
r.squaredGLMM(seedcount2)
```

```{r}
seedcount2 |>
  as_flextable()
```

Visualizing:
```{r}
predictions <- ggpredict(seedcount2, terms = c("total_nr_infl", "plant_nr", "tag_num")) %>% 
  rename(total_nr_infl = group,
        plant_nr = facet)

ggplot(seedcount, aes(x = total_nr_infl, y = plant_nr, fill = plant_nr)) +
  geom_point(aes(color = plant_nr), alpha = 0.5) +
  facet_wrap(~tag_num, scales = "free_y") +
  geom_line(data = predictions, aes(x = x, y = predicted, color = plant_nr)) +
  geom_ribbon(data = predictions, aes(x = x, y = predicted, ymin = conf.low, ymax = conf.high, fill = plant_nr), alpha = 0.2) +
  scale_fill_manual(values = c("yes" = "blue", "no" = "orange")) +
  scale_color_manual(values = c("yes" = "blue", "no" = "orange")) +
  theme_bw() +
  facet_wrap(~tag_num, scales = "free_y") +
  labs(fill = "plant_nr", color = "plant_nr")
```

