---
title: "ENVS-193DS_final"
author: "Lauren Stiles"
date: 06--2023
format:
  html:
    toc: true
    toc-location: left
    code-fold: true
    theme: yeti
execute: 
  message: false
  warning: false 
---

# notes about what to include -- 1. Context: for example, in biological terms, what is the question? Hypothesis/hypotheses?
What are the main results? What is your interpretation of those results?
2. Logic: for example, how did you process the data? What steps did you go through to get
your result? What assumptions did you check, and how? What test did you run and why?
What tools (e.g. packages, functions) did you use?
3. Transparency: for example, are your code and data organized?Is your code annotated and
spaced out?Is it publicly available? Did you write an informative README?
4. Clarity: for example, is your writing conveying your thoughts appropriately?Is your
document clean and organized? Are sections clear?
5. Communication: for example, are your figures and tables clear? Have you used colors,
shapes, or other aesthetics to create a clean, visually appealing plot? Have you used the
appropriate plot for your data set and analysis? Have you rounded/summarized numbers
to the appropriate number of digits?
#load libraries: 
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(here) 
library(janitor) 
library(readxl)
library(naniar)
library(car)
```

## **Problem 1**

# Question: How does total seed number differ between kangaroo rat mound locations?

# Introduction: 3-5 citations/sentences #DO THIS 

# Methods 
data available [here](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-sev.208.102459)
We read in the data and cleaned it by selecting the columns we are interested in and renaming the locations (base of mound, edge of grass, grass interspace) for more straightforward interpretation. We chose to use an ANOVA test to compare the count sample means since this kind of test can compare means between more than two groups. We tested the null hypothesis that each of mean seed counts from the different mound locations are all the same. The alternative hypothesis is that at least one of the means is different than the other sample means. We tested the assumption that the samples have a normal distribution and equal variances. The qqplot shows that the data is mostly normal, since most of the facets represent the data as a straight line. There are deviations from normality from the data in each, especially in the base of mound group. But, ANOVA is robust even when this assumption is not met, so we decided to continue with the analysis. We used a Levene test to determine whether the data has equal variances. The null hypothesis of this test is that the variances are equal and the alternative hypothesis is that the variances are not equal. The p value from this test is <0.05, so we reject the null. It does not seem like the variances are equal. Again, since ANOVA is robust, we continued with this test.  #EDIT THIS A LOT MORE 

Read in data: 
```{r message = FALSE, warning = FALSE}
seed_dat <- read_csv(here("data/sev208_kratseedbank_20120213.csv"))
```

Clean data: 
```{r}
#create new data frame with kangaroo rat mound locations and seed numbers found 
seed_dat_clean <- seed_dat |> 
                  #select columns of interest
                  select(mnd, loc, seeds) |> 
                  #remove "D" from loc column becuase there is no data defintion on the EDI portal 
                  filter(loc != "D") |>
                  #rename columns 
                  mutate(loc = case_when(loc == "B" ~ "Base of mound", 
                                          loc == "E" ~ "Edge of grass", 
                                          loc == "I" ~ "Grass interspace"))  
```


Check Assumptions: 
```{r warning = FALSE, message = FALSE}
#first, checking for missing data 
gg_miss_var(seed_dat_clean) #no missing data 

#check for normal distribution
ggplot(data = seed_dat_clean, aes(sample = seeds)) + 
  #qqplot with dots 
  stat_qq() + 
  #add line to qqplot
  stat_qq_line(color = "red") + 
  #split plot into calcium-treated and reference watersheds 
  facet_wrap(~loc) 
#looks more or less normally distributed -- pretty linear 

#check for equal variance of seed count data using a Levene test
seed_var_test <-  leveneTest(seeds ~ loc, data = seed_dat_clean, var.equal = TRUE)
seed_var_test
```

Run ANOVA: 
```{r}
aov(seeds ~ loc, data = seed_dat_clean)

```

Visualization: 
```{r}

```

Table: 
```{r}

```

## **Bibliography**
Koontz, T. and H. Simpson. 2013. Effects of Kangaroo Rat Mounds on Seed Banks of Grass
and Shrublands at the Sevilleta National Wildlife Refuge, New Mexico (2001) ver 102459.
Environmental Data Initiative.

Koontz and Simpson 2010, “The composition of seed banks on kangaroo rat
(Dipodomysspectabilis) mounds in a Chihuahuan Desert grassland.” Journal of Arid Environments.
https://doi.org/10.1016/j.jaridenv.2010.03.008

